---
title: "protein_helen"
author: "Doug McNeall"
date: "2020-07-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Classification of protein tail-anchoring locations.  

This code was use for the analysis in the paper "Predicting the targeting of tail-anchored proteins to subcellular compartments in mammalian cells" by Costello et al. (2017). Journal of Cell Science 2017 130: 1675-1687; doi: 10.1242/jcs.200204  

<https://jcs.biologists.org/content/130/9/1675.short>

Code by Doug McNeall based on data by Joe Costello.

The code currently uses a support vector machine (SVM) for classification of tail-anchoring location, depending on TMD hydrophobicity and tail charge.

Code is here: <https://github.com/dougmcneall/protein/blob/master/docs/protein.Rmd>
  
## Setup
```{r}
set.seed(42)

library(RColorBrewer)
library(e1071)
library(MASS)

paired <- brewer.pal(6,'Paired') 
blues <- brewer.pal(9,'Blues')

mitoblue <- '#A2D9F7'
pogreen  <- '#A8D4AF'
ersalmon <- '#F6B096'

catpat <- paired
catpat[1] <- ersalmon
catpat[2] <- mitoblue
catpat[6] <- pogreen
catpat[3] <- paired[6]
catpat[4] <- paired[2]
catpat[5] <- paired[4]

truncpat <- c(ersalmon, mitoblue, pogreen)

```


## Data

```{r}
protein <- read.csv('../data/protein.csv')
# Or, we could just load up the truncated file (added 10th December 2015)
# This breaks the colour palettes at the moment.
protein_trunc <- read.csv('../data/protein_trunc.csv')

# factors are alphabetical, so as.numeric gives
# ER          = 1
# MIT0        = 2
# MITO/ER     = 3
# MITO/PO     = 4
# MITO/PO/ER  = 5
# PO          = 6

# Training data 
train <- subset(protein, select = c('tmd_gravy', 'tail_charge'))
cl <- protein$location

# Data truncated to just the main 3 classes

#protein_trunc <- subset(protein, location=='ER' | location=='MITO' | location=='PO')

train_trunc <- subset(protein_trunc, select = c('tmd_gravy', 'tail_charge'))
cl_trunc <- protein_trunc$location

upper <- apply(train_trunc, 2, max)
lower <- apply(train_trunc, 2, min)

tmd_gravy_seq <- seq(from=lower[1], to=upper[1], length.out = 50)
tail_charge_seq <- seq(from=lower[2], to=upper[2], length.out = 50)

# A grid of test points that covers the entire data space
test <- as.matrix(expand.grid(tmd_gravy_seq, tail_charge_seq))
colnames(test) <- colnames(train_trunc)

# skip the last row, as there is no tail charge 
target<- read.table('../data/protein_target.txt', sep = '\t', nrows = 244, header = TRUE)
colnames(target) <- colnames(train_trunc)

mutants <- read.csv('../data/mutants.csv')

# A little aside to get the factors in the data sets to be the same. rbind()
# changes the as.numeric factors to match. 
combined <- rbind(protein, mutants)

mutants.ix <- seq(from = (nrow(protein)+1), to = nrow(protein)+nrow(mutants), by = 1)
mutants_refactor <- combined[mutants.ix, ]

# "Shared location" data
protein_shared <- subset(protein, location=='MITO/ER' | location=='MITO/PO' | location=='MITO/PO/ER')


# some new test sites

test.sites <- data.frame(cbind(c(1.54, 1.43, 2.34), c(3.9, 4.6, -0.1)))

colnames(test.sites) <- c('tmd_gravy', 'tail_charge')

#1) TMD GRAVY = 1.54 Charge =  3.9
#2) TMD GRAVY = 1.43 Charge =  4.6
#3) TMD GRAVY = 2.34 Charge = -0.1

```

## Plot all the data (with mutants)

```{r, fig.width = 7, fig.height = 7}

par(las = 1, cex = 1.3)
plot(protein$tmd_gravy, protein$tail_charge, bg = catpat[protein$location],
 pch = 21, col = 'black',
 xlab = 'tmd_gravy', ylab = 'tail_charge')

points(mutants_refactor$tmd_gravy, mutants_refactor$tail_charge, col = 'black', bg = catpat[mutants_refactor$location], pch = 24)

legend('topleft', legend = unique(protein$location), pt.bg = catpat[unique(protein$location)], pch = 21, text.col = catpat[unique(protein$location)], cex = 0.8, bty = 'n' )
legend('top', legend = c( 'triangles indicate mutants'), pch = c(24), cex = 0.8, bty = 'n')
#dev.print(pdf, file='protein_and_mutants.pdf', width = 6, height = 6)

```

```{r, fig.width = 7, fig.height = 7}

# Data truncated to just the 3 main classes
#dev.new()
par(las = 1, cex = 1.3)
plot(protein_trunc$tmd_gravy, protein_trunc$tail_charge, col = 'black', bg = truncpat[protein_trunc$location], pch = 21,
     xlab = "tmd_gravy", ylab = "tail_charge")

legend('topleft', legend = unique(protein_trunc$location), col = 'black', pt.bg = truncpat[unique(protein_trunc$location)], pch = 21, text.col = truncpat[unique(protein_trunc$location)] )
#dev.print(pdf, file='protein_trunc.pdf', width = 6, height = 6)


```

## Support Vector Machine Classification  

Train a non-probabilistic support vector machine (SVM) on the truncated data, and visualise the decision boundaries.

```{r, fig.width = 7, fig.height = 7}

# Non-probabilistic SVM fit
svm.fit <- svm(location~., data = protein_trunc, probability = FALSE)
svm.pred <- predict(svm.fit, newdata=test, probability = FALSE)

# misclassification rate for the svm is ~18%
svm.mcr = 1 - (sum(predict(svm.fit)==protein_trunc$location)/ length(protein_trunc$location))

par(las = 1, cex = 1.3)
plot(test, col = truncpat[svm.pred], pch = 20, cex = 0.8, main = 'SVM classifier prediction')
points(protein_trunc$tmd_gravy, protein_trunc$tail_charge, col = 'black', bg = truncpat[protein_trunc$location], pch = 21)

legend('topleft', legend = unique(protein_trunc$location),
pt.bg = truncpat[unique(protein_trunc$location)],
pch = 21,
col = 'black',
text.col=truncpat[unique(protein_trunc$location)],
 bg = 'white')
#dev.print(pdf, file = 'svm_prediction.pdf', width = 6, height = 6)
 
```

## E.cuniculi non-probabilistic prediction
```{r,fig.width = 7, fig.height = 7}

E.cunic <- read.csv('../data/helen/E.cuniculi.csv')

svm.pred.E.cunic <- predict(svm.fit, newdata=E.cunic, probability = FALSE)

mx <- apply(E.cunic, 2, max)
mn <- apply(E.cunic, 2, min)

par(las = 1, cex = 1.3)
plot(test, col = truncpat[svm.pred], pch = 20, cex = 0.8, main = 'SVM classifier prediction',
     xlim = c(mn[1], mx[1]),
     ylim = c(mn[2], mx[2])
     )
points(protein_trunc$tmd_gravy, protein_trunc$tail_charge, col = 'black', bg = truncpat[protein_trunc$location], pch = 21)

points(E.cunic$tmd_gravy, E.cunic$tail_charge, col = 'red', bg = truncpat[svm.pred.E.cunic], pch = 21)

legend('topleft', legend = unique(protein_trunc$location),
pt.bg = truncpat[unique(protein_trunc$location)],
pch = 21,
col = 'black',
text.col=truncpat[unique(protein_trunc$location)],
 bg = 'white')
#dev.print(pdf, file = 'svm_prediction.pdf', width = 6, height = 6)


```


```{r}

# create a list of all the csv files in the directory
filelist <- dir('../data/helen/', pattern = '*.csv')

#loop over the files
for(i in filelist){
  # put all the steps of the previous code chunk in here, e.g.
  
  # read data for prediction locations
  dat <- read.csv(paste0('../data/helen/', i))
  # make the predictions
  newpred <- predict(svm.fit, newdata = dat, probability = FALSE)
  
  # plot here
  # ...
  
}

```


##Probabilistic SVM fit

Train a probabilistic support vector machine (SVM) on the truncated data, and visualise the probability contours.

```{r, fig.width = 7, fig.height  = 7}
# Why does this predict some non-seen classes?
svm.fit.prob <- svm(location~., data = protein_trunc, probability = TRUE)

svm.pred.prob <- predict(svm.fit.prob, newdata=test, probability = TRUE)

# How many does the svm fit correctly?

fit.hit <- sum(svm.fit.prob$fitted == protein_trunc$location)
fit.miss <- length(protein_trunc$location) - fit.hit
prob.mcr <- fit.miss / length(protein_trunc$location) 

svm.prob.df <- attr(svm.pred.prob,'prob')
mitomat <- matrix(svm.prob.df[,'MITO'], nrow = length(tmd_gravy_seq))
pomat <- matrix(svm.prob.df[,'PO'], nrow = length(tmd_gravy_seq))
ermat <- matrix(svm.prob.df[,'ER'], nrow = length(tmd_gravy_seq))


protein_trunc.fit <- cbind(protein_trunc,as.character(svm.fit.prob$fitted))
colnames(protein_trunc.fit)[4] <- c('predicted')
#write.matrix(protein_trunc.fit, file = 'protein_trunc_fit.txt')


n <- length(protein_trunc$location)
cv.out <- rep(NA, n)
# A simple looping leave-one-out cross validation
for(i in 1:n){
	
	cv.fit <-  svm(location~., data = protein_trunc[-i, ], probability = TRUE)
	cv.pred <- predict(cv.fit, newdata=protein_trunc[i, ], probability = TRUE)
	cv.out[i] <- as.character(cv.pred) 

}

loo.hitrate <- sum(cv.out == as.character(protein_trunc$location)) / length(protein_trunc$location)

labcex = 0.8
pt.cex = 1.3

par(las=1, mar = c(4,4,1,1))
plot(test, col = truncpat[svm.pred], type = 'n', main = '',
xlab = 'TMD GRAVY', ylab = 'Tail charge')#, main = 'SVM classifier class probability')

contour(tmd_gravy_seq, tail_charge_seq, mitomat,
	add = TRUE, levels = c(0.5,0.6,0.7,0.8,1), col = mitoblue,labcex = labcex)
contour(tmd_gravy_seq, tail_charge_seq, pomat,
	add = TRUE, levels = c(0.5,0.6,0.7,0.8,1), col = pogreen, labcex = labcex)
contour(tmd_gravy_seq, tail_charge_seq, ermat,
	add = TRUE,levels = c(0.5,0.6,0.7,0.8), col = ersalmon, labcex = labcex)
    
points(protein_trunc$tmd_gravy, protein_trunc$tail_charge,
	col = 'black',bg = truncpat[protein_trunc$location], pch = 21, cex = pt.cex)
points(test.sites$tmd_gravy, test.sites$tail_charge,
	col = 'black', pch = 21, lwd = 1.5, cex = pt.cex)

legend('topleft', legend = c(as.character(unique(protein_trunc$location)), 'test'),
	col = 'black', pt.bg = c(truncpat[unique(protein_trunc$location)],'white'), pch = 21, text.col = c(truncpat[unique(protein_trunc$location)], 'black'), bg = 'white', pt.lwd = c(rep(1,3),1.5), pt.cex = pt.cex)

#dev.print(pdf, file='svm_probability_contour.pdf', width = 5, height = 5)
#dev.print(tiff, file='svm_probability_contour.tif', width = 480, height = 480)


svm.protein <- svm(location~., data = protein, probability = TRUE)#!! 
protein.hit <- sum(svm.protein$fitted == protein$location)

protein.fit <- cbind(protein,as.character(svm.protein$fitted))
colnames(protein.fit)[4] <- c('predicted')

#write.matrix(protein.fit, file = 'protein_fit.txt')
```


# SVM classifier target predictions

Predict tail anchoring location at new values of tmd_gravy and tail_charge.

```{r, fig.width = 7, fig.height = 7}
# ---------------------------------------------------------
# Predict the location in the target data
# ---------------------------------------------------------

target.prob <- predict(svm.fit.prob, newdata = target, probability = TRUE )

par(las=1)
plot(target, type = 'n',  main = 'SVM classifier target predictions')
points(target, col = truncpat[target.prob], pch = 21)
contour(tmd_gravy_seq, tail_charge_seq, mitomat, add = TRUE, levels = c(0.5,0.6,0.7,0.8,1), col = mitoblue)
contour(tmd_gravy_seq, tail_charge_seq, pomat,add = TRUE, levels = c(0.5,0.6,0.7,0.8,1), col = pogreen)
contour(tmd_gravy_seq, tail_charge_seq, ermat,add = TRUE, levels = c(0.5,0.6,0.7,0.8), col = ersalmon)
points(protein_trunc$tmd_gravy, protein_trunc$tail_charge, col = 'black',bg = truncpat[protein_trunc$location], pch = 21)
legend('topright', legend = unique(protein_trunc$location), col = 'black', pt.bg = truncpat[unique(protein_trunc$location)], pch = 21, text.col = truncpat[unique(protein_trunc$location)], bg = 'white')

#dev.print(pdf, file = 'protein_target.pdf', width = 6, height = 6)


#write.matrix(cbind(target,attr(target.prob, 'probabilities'),target.prob), file = 'target_predictions.txt')

```


## Probabilistic fit for E.cuniculi
```{r, fig.width = 7, fig.height = 7}

target.prob.E.cunic <- predict(svm.fit.prob, newdata = E.cunic, probability = TRUE )

par(las=1)
plot(E.cunic, type = 'n',  main = 'SVM classifier target predictions')
#points(target, col = truncpat[target.prob], pch = 21)
points(E.cunic, col = truncpat[target.prob.E.cunic], pch = 21)

contour(tmd_gravy_seq, tail_charge_seq, mitomat, add = TRUE, levels = c(0.5,0.6,0.7,0.8,1), col = mitoblue)
contour(tmd_gravy_seq, tail_charge_seq, pomat,add = TRUE, levels = c(0.5,0.6,0.7,0.8,1), col = pogreen)
contour(tmd_gravy_seq, tail_charge_seq, ermat,add = TRUE, levels = c(0.5,0.6,0.7,0.8), col = ersalmon)

points(protein_trunc$tmd_gravy, protein_trunc$tail_charge, col = 'black',bg = truncpat[protein_trunc$location], pch = 21)

#points(E.cunic$tmd_gravy E.cunic$tail_charge, col = 'black',bg = truncpat[protein_trunc$location], pch = 21)

legend('topright', legend = unique(protein_trunc$location), col = 'black', pt.bg = truncpat[unique(protein_trunc$location)], pch = 21, text.col = truncpat[unique(protein_trunc$location)], bg = 'white')


```








